\section{IO and Builtin Values and Predicates}

While not this material is neither novel nor particularly difficult, 
I'm including this section as both a specification and a guide 
to future implementors of logic programming languages with proof search.

The first thing to note when programming in Caledon is that searching for items of type $\m{prop}$ 
can might IO, and thus IO can be performed during typechecking.  
This can be understood as the set of axioms differing based on the environment available.

IO is performed when the evaluation function encounters a query for a builtin IO performing function.

\begin{lstlisting}
eval : Proposition -> Environment Formula
eval (a %*$\in$ *) ``print'' Str) = ( if gvar a then print str else ()
                                    ; return (a %* $\doteq $ *) printImp Str)
                                    )
\end{lstlisting}

It is important to include the check that $a$ has not already been resolved such that repeated io actions are not performed
when nondeterministically proof searching.  

It is tempting to define predicates that take input as taking as an argument 
a function that uses the input.  This is in fact a valid way to define such functions, 
and permits for hints of directionality in the types.  
However, it is still possible to escape.  

\begin{lstlisting}
builtin readLine : (string -> prop) -> prop

defn readLineEscape : string -> prop
  as \ s : string . readLine (\ t . t =:= s )

\end{lstlisting}

Ensuring variables do not escape their intended scope is necessary to ensuring
that the intended IO action is only executed once, and not multiple times during proof search.

While it is in fact possible to reason about nondeterministic IO, it is desirable to also have actions
that can not be executed twice, for which nondeterminism is not possible.

For this, the notion of an monad is in fact usefull.  
In this setting io is presented as a series of builtin axioms.

\begin{lstlisting}
defn io : prop -> prop
   | bind     = io A -> (A -> io B) -> io B
   | return   = A -> io A
   | ioReadLine = io string
   | ioPrint    = string -> io unit
\end{lstlisting}

We can now no longer write $A \in \m{readLine}$ though since $\m{readLine} : \m{prop}$ 
is a value and thus has no inhabitants. 

Instead, an interpretation predicate can be created which maps these dummy io actions to real actions.

\begin{lstlisting}
defn run : io A -> A -> prop
  >| runBind = run (bind IOA F) V
            <- run IOA A 
            <- run (F A) V 
  >| runReturn   = run (return V) V
  >| runReadLine = run ioReadLine A <- readLineEscape A
  >| runPrint    = run (ioPrint S) one <- print S
\end{lstlisting}

It is important to note that since the type system is the calculus of constructions, io actions constructed
from $io$ will be total, severely limiting their utility.
More complex io actions and interpreters can be generated, most importantly ones involving recursion 
or infinite loops.
